PineAPPL:\\ combining EW and QCD corrections for PDF fits

### 2020-08-09 paper/sec-doublecounting.tex

\begin{figure}
    \centering
    \begin{overpic}[width=0.6\textwidth, trim=0.cm 11cm 0.cm 10cm, clip=True]{figures/dy-pi}
        \put (5, 33) {\large $\mathrm{q}$}
        \put (20, 37) {\large $\mathrm{q}$}
        \put (38, 5) {\large $\bar{\ell}$}
        \put (38, 31) {\large $\ell$}
        \put (11, 6) {\large $\gamma$}
        %
        \put (65, 9) {\large $\mathrm{q}$}
        \put (73, 30) {\large $\mathrm{q}$}
        \put (96, 25) {\large $\ell$}
        \put (96, 8) {\large $\bar{\ell}$}
        \put (55, 26) {\large $\gamma$}
    \end{overpic}
    \caption{\label{fig:dy-pi}
    Photon-induced (left) and quark-induced (right) contributions to the Drell-Yan process. In black, the LO process is shown.
    In red, the initial-state splitting leading to the real-emission $\mathrm{q} \gamma \to \ell \bar{\ell} \mathrm{q}$ is highlighted. Such a
    real emission enters in the NLO EW corrections.}
\end{figure}
An example is the subtraction of (irreducible) background processes which must not be considered as such. A very blatant case
is neutral-current Drell--Yan, where the signal process is the production of an opposite-sign lepton pair, which starts
at $\mathcal O(\alpha^2)$. Because this process is usually thought
as a quark-initiated $s$-channel mechanism ($\mathrm{q} \bar{\mathrm{q}} \to \gamma^*/\mathrm{Z} \to \ell \bar{\ell}$), in many analyses the photon-induced component,
$\gamma \gamma \to \ell \bar{\ell}$ in the $t$ channel, is considered to be a different process, and therefore as a background and subtracted.
The subtraction from the measured data is done by calculating the theoretical predictions of the double-photon initiated contribution, possibly including (ill-defined) higher-order
corrections. For example, in refs.~\cite{Aaboud:2017ffb,Aad:2016zzw} (a similar statement appears also in an older analysis~\cite{Aad:2013iua}), one reads:
\begin{quote}
The photon-induced process, $\gamma\gamma \to \ell \bar{\ell}$, is simulated at LO using Pythia 8
and the MRST2004qed PDF set~\cite{Martin:2004dh}. The expected yield for this process also accounts for 
NLO QED/EW corrections from references~\cite{Bardin:2012jk,Bondarenko:2013nu}, which decrease the yield by approximately \SI{30}{\percent}.
\end{quote}
Such a distinction, which is unphysical and incorrect in quantum mechanics, may be somehow justified at LO. Beyond this order, it is simply wrong.
Indeed, at $\mathcal O(\alpha^3)$, the reaction $\mathrm{q} \gamma \to \ell \bar{\ell} \mathrm{q}$ becomes possible, which
includes both kind of topologies discussed above, and needs both in order to yield an IR-finite result, see Fig.~\ref{fig:dy-pi} (as a consequence, one cannot speak of EW corrections to $\gamma \gamma \to \ell \bar{\ell}$). While it is common sense that the QCD counterpart of this subtraction should never 
be performed --- no one would
consider to \enquote{subtract} the gluon-initiated contribution to top-pair production in top analyses --- seemingly it is not so
when EW corrections are considered.

%%{\bf It looks like CMS does things in a pretty consistent manner~\cite{Sirunyan:2018owv, CMS:2014jea}. Shall we mention? We should
%%not make ATLAS appear as the bad guys, and CMS as the good ones.}

A second example is related to removing EW effects from data. These can be either the full EW corrections
or just a part of them. In either case, a comparison between these data and a NLO-EW accurate simulation aimed at the extraction of some parameter would be meaningless, as some effects included in the latter
have been removed from the former. The typical example relevant for LHC is the deconvolution of effects due to multiple-photon radiation
from light particles in the final state. This applies mostly
to processes such as neutral- or charge-current Drell--Yan, especially when electrons are considered. The problem lies in the fact that
 electrons, and to a lesser extent muons, tend to radiate photons, which are not accounted
for in QCD-only matrix-elements. Thus, leptons that are measured in the detector are less energetic, and this fact is compensated for
by inverting a photon shower. The resulting dataset is e.g.\ referred to as \emph{pre-FSR} with observables defined in terms of \emph{Born-level electrons} (see e.g.\ ref.~\cite{Aad:2015auj} for its definition).
These datasets are needed for and correctly used in QCD-only PDF determinations, since the EW corrections to some DY observables can be significant, and excluding them would therefore degrade the quality of the fit.
However, in fits including fixed-order EW corrections the problem with this definition is that the first photon emission is included exactly at the matrix-element level. The inclusion of
subsequent emission would require the matching with the QED shower, which is not yet available for general processes.

>>>>

\begin{figure}
    \centering
    \begin{overpic}[width=0.6\textwidth, trim=0.cm 11cm 0.cm 10cm, clip=True]{figures/dy-pi}
        \put (5, 33) {\large $\mathrm{q}$}
        \put (20, 37) {\large $\mathrm{q}$}
        \put (38, 5) {\large $\bar{\ell}$}
        \put (38, 31) {\large $\ell$}
        \put (11, 6) {\large $\gamma$}
        %
        \put (65, 9) {\large $\mathrm{q}$}
        \put (73, 30) {\large $\mathrm{q}$}
        \put (96, 25) {\large $\ell$}
        \put (96, 8) {\large $\bar{\ell}$}
        \put (55, 26) {\large $\gamma$}
    \end{overpic}
    \caption{\label{fig:dy-pi}
    Photon-induced (left) and quark-induced (right) contributions to the Drell-Yan process. In black, the LO process is shown.
    In red, the initial-state splitting leading to the real-emission $\mathrm{q} \gamma \to \ell \bar{\ell} \mathrm{q}$ is highlighted. Such a
    real emission enters in the NLO EW corrections.}
\end{figure}
An example is the subtraction of (irreducible) background processes which must not be considered as such. A very blatant case
is neutral-current Drell--Yan, where the signal process is the production of an opposite-sign lepton pair, which starts
at $\mathcal O(\alpha^2)$. Because this process is usually thought
as a quark-initiated $s$-channel mechanism ($\mathrm{q} \bar{\mathrm{q}} \to \gamma^*/\mathrm{Z} \to \ell \bar{\ell}$), in many analyses the photon-induced component,
$\gamma \gamma \to \ell \bar{\ell}$ in the $t$ channel, is considered a different process, and therefore as a background and subtracted.
The subtraction from the measured data is done by calculating the theoretical predictions of the double-photon initiated contribution, possibly including (ill-defined) higher-order
corrections. For example, in refs.~\cite{Aaboud:2017ffb,Aad:2016zzw} (a similar statement appears also in an older analysis~\cite{Aad:2013iua}), one reads:
\begin{quote}
The photon-induced process, $\gamma\gamma \to \ell \bar{\ell}$, is simulated at LO using Pythia 8
and the MRST2004qed PDF set~\cite{Martin:2004dh}. The expected yield for this process also accounts for 
NLO QED/EW corrections from references~\cite{Bardin:2012jk,Bondarenko:2013nu}, which decrease the yield by approximately \SI{30}{\percent}.
\end{quote}
Such a distinction, which is unphysical and incorrect in quantum mechanics, may be somehow justified at LO. Beyond this order, it is simply wrong.
Indeed, at $\mathcal O(\alpha^3)$, the reaction $\mathrm{q} \gamma \to \ell \bar{\ell} \mathrm{q}$ becomes possible, which
includes both kind of topologies discussed above, and needs both in order to yield an IR-finite result, see Fig.~\ref{fig:dy-pi} (as a consequence, one cannot speak of EW corrections to $\gamma \gamma \to \ell \bar{\ell}$). While it is common sense that the QCD counterpart of this subtraction should never 
be performed --- no one would
consider to \enquote{subtract} the gluon-initiated contribution to top-pair production in top analyses --- seemingly it is not so
when EW corrections are considered.

A second example is related to removing EW effects from data. These can be either the full EW corrections
or just a part of them. In either case, a comparison between these data and a NLO-EW accurate simulation aimed at the extraction of some parameter would be meaningless, as some effects included in the latter
have been removed from the former. The typical example relevant for LHC is the deconvolution of effects due to multiple-photon radiation
from light particles in the final state. This applies mostly
to processes such as neutral- or charge-current Drell--Yan, especially when electrons are considered. The problem lies in the fact that
 electrons, and to a lesser extent muons, tend to radiate photons, which are not accounted
for in QCD-only matrix elements. Thus, leptons that are measured in the detector are less energetic, and this fact is compensated for
by inverting a photon shower. The resulting dataset is e.g.\ referred to as \emph{pre-FSR} with observables defined in terms of \emph{Born-level electrons} (see e.g.\ ref.~\cite{Aad:2015auj} for its definition).
These datasets are needed for and correctly used in QCD-only PDF determinations, since the EW corrections to some DY observables can be significant, and excluding them would therefore degrade the quality of the fit.
In fits including fixed-order EW corrections the problem with this definition, besides double counting, is that the first photon emission is included exactly at the matrix-element level. The inclusion of
subsequent emission would require the matching with the QED shower, which is not yet available for general processes.

### 2020-08-09 paper/sec-pineappl.tex

\section{PDF-independent storage of phase-space weights with \texorpdfstring{\textsc{PineAPPL}}{PineAPPL}}
\label{sec:pineappl}

@@

Fixed-order partonic cross sections $a + b \to X$ supported by \textsc{PineAPPL} are written as an expansion in powers of the strong coupling $\alphas$, the electromagnetic coupling $\alpha$, and the logarithms of $\xi_\mathrm{R} = \mu_\mathrm{R} / Q$ and $\xi_\mathrm{F} = \mu_\mathrm{F} / Q$,
\begin{multline}
\frac{\mathrm{d} \sigma_{ab}}{\mathrm{d} \mathcal{O}} (x_1, x_2, \mathcal{O}, \xi_\mathrm{R}, \xi_\mathrm{F}) \\
= \sum_{k,l,m,n} \alphas^k \left( \xi_\mathrm{R}^2 Q^2 \right) \alpha^l \log^m ( \xi_\mathrm{R}^2 ) \log^n ( \xi_\mathrm{F}^2 ) W_{ab}^{(k,l,m,n)} \left( x_1, x_2, Q^2, \mathcal{O} \right) \text{.}
\label{eq:expansion}
\end{multline}
The cross section shown above is differential w.r.t.\ the observable $\mathcal{O}$, which, in general, is a function of phase space and subject to the usual conditions (soft and collinear safety, etc.).

>>>>

\section{PDF-independent storage of phase-space weights with PineAPPL}
\label{sec:pineappl}

@@

Fixed-order partonic cross sections $a + b \to X$ supported by \textsc{PineAPPL} are written as an expansion in powers of the strong coupling $\alphas$, the electromagnetic coupling $\alpha$, and the logarithms of $\xi_\mathrm{R} = \mu_\mathrm{R} / Q$ and $\xi_\mathrm{F} = \mu_\mathrm{F} / Q$,
\begin{multline}
\frac{\mathrm{d} \sigma_{ab}}{\mathrm{d} \mathcal{O}} (x_1, x_2, \mathcal{O}, \xi_\mathrm{R}, \xi_\mathrm{F}) \\
= \sum_{k,l,m,n} \alphas^k \left( \xi_\mathrm{R}^2 Q^2 \right) \alpha^l \log^m ( \xi_\mathrm{R}^2 ) \log^n ( \xi_\mathrm{F}^2 ) W_{ab}^{(k,l,m,n)} \left( x_1, x_2, Q^2, \mathcal{O} \right) \text{.}
\label{eq:expansion}
\end{multline}
The cross section shown above is differential with respect to the observable $\mathcal{O}$, which, in general, is a function of phase space and subject to the usual conditions (soft and collinear safety, etc.).
